A Formalization of Typed Lua

### 2015-06-13 doc/papers/paper/paper.tex
DLS: some fixes to the section on projection types

\begin{figure*}[t]
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IfFilterProjThen}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
	finil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjElse}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\
	 S_e = finilpt(\Pi(x), i)\\
	fonil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjBoth}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
	\env_{4} = join(\env_{2}, \env_{3}) 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\end{array}
\]
\caption{Rules for Projection Types}
\end{figure*}

@@

Rule {\sc T-ExpListProj} is the typing rule for this auxiliary
relation. It does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type.

@@

Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
filter the union referenced by $\pi_{i}^{x}$, respectively
filtering $\Nil$ out at every i-th component (removing
the entire tuple if it is $\Nil$) and filtering $\Nil$ in at every i-th component (removing the entire tuple if it is not $\Nil$).

@@

Assignment to an l-value with a projection type is forbidden. Allowing assignment
would be unsound, as it could break the dependency among the
components of the union, as the following example shows:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}

The projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$.

>>>>

\begin{figure*}[t]
\label{fig:projtypes}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-LocalProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\begin{array}{c}
\mylabel{T-ExpListProj}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|) \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x}{*}, \env_{f}, (x,S)}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IfFilterProjThen}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i)\\ 
	finil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjElse}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\
	 S_e = finilpt(\Pi(x), i)\\
	fonil(proj(\Pi(x), i)) = {\bf void} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{2} 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-IfFilterProjBoth}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \pi_{i}^{x} \\ S_t = fonilpt(\Pi(x), i) \; S_e = finilpt(\Pi(x), i)\\
	\env_{1}, \penv[x \mapsto S_t] \vdash s_{1}, \env_{2} \\
	\env_{1}, \penv[x \mapsto S_e] \vdash s_{2}, \env_{3} \\
	\env_{4} = join(\env_{2}, \env_{3}) 
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}\\\\
%\begin{array}{rcll}
%last(F_1{*} \sqcup F_2{*}) & = & F_1{*} & \mathrm{if} \; F_2{*} \subtype F_1{*}\\
%last(F_1{*} \sqcup F_2{*}) & = & F_2{*} & \mathrm{if} \; F_1{*} \subtype F_2{*}\\
%last(F_1{*} \sqcup F_2{*}) & = & (F_1 \cup F_2){*} & \mathrm{otherwise}\\
%last(F{*} \sqcup T \times P) & = & last(\Nil{*} \sqcup P) \\
%last(T \times P \sqcup F{*}) & = & last(P \sqcup \Nil{*}) \\
%last(P \sqcup S) & = & last(P \sqcup last(S)) \\
%last(S \sqcup P) & = & last(last(S) \sqcup P) \\
%last(S_1 \sqcup S_2) & = & last(last(S_1) \sqcup last(S_2))
%\end{array}\\\\
\begin{array}{rcll}
proj(S_1 \sqcup S_2, i) & = & proj(S_1, i) \cup proj(S_2, i) \\
proj(F{*}, i) & = & F \cup \Nil \\
proj(T \times P, 1) & = & T \\
proj(T \times P, i) & = & proj(P, i-1)
\end{array}\\\\
\begin{array}{rcll}
fonilpt(P_1 \sqcup P_2) & = & P_2 & \mathrm{if} \; fonil(proj(P_1, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 & \mathrm{if} \; fonil(proj(P_2, i)) = {\bf void} \\
fonilpt(P_1 \sqcup P_2) & = & P_1 \sqcup P_2 & \mathrm{otherwise}\\
fonilpt(P \sqcup S) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(P \sqcup S) & = & P \sqcup fonilpt(S) & \mathrm{otherwise}\\
fonilpt(S \sqcup P_2) & = & fonilpt(S) & \mathrm{if} \; fonil(proj(P, i)) = {\bf void} \\
fonilpt(S \sqcup P) & = & fonilpt(S) \sqcup P & \mathrm{otherwise}\\
fonilpt(S_1 \sqcup S_2) & = & fonilpt(S_1) \sqcup fonilpt(S_2)
\end{array}\\\\
\begin{array}{rcll}
fonil(F_1 \cup F_2) & = & fonil(F_1) \cup fonil(F_2) \\
fonil(F) & = & {\bf void} & \mathrm{if} \; F \subtype \Nil \; \mathrm{and} \; \Nil \subtype F\\
fonil(F) & = & F & \mathrm{otherwise}
\end{array}
\end{array}
\]
\caption{Rules for Projection Types}
\end{figure*}

@@

Rule {\sc T-ExpListProj} is the typing rule for this auxiliary
relation. It does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type.
The rule generates as many projection types as the the length
of the longest tuple.

@@

Rule \textsc{T-IfFilterNilProj} uses the auxiliary functions \emph{fonilpt} ({\em filter-out nil in projection type}) and \emph{finilpt} ({\em filter-in nil in projection type}) to
filter the union referenced by $\pi_{i}^{x}$, respectively
filtering $\Nil$ out at every i-th component (removing
the entire tuple if it is $\Nil$) and filtering $\Nil$ in at every i-th component (removing the entire tuple if it is not $\Nil$). These functions respectively use the functions $fonil$ and $finil$ to filter regular unions, plus $proj$ to test projected components. The definitions of $finilpt$ and $finil$ are analogous to the definitions of $finilpt$ and $finil$.

@@

Assignment to an l-value with a projection type is forbidden. Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
the projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}
